---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file, then
     regenerate with devtools::build_readme(). Do not use knitr::knit(): it
     processes the code but leaves this YAML header in the output as literal
     text, which GitHub and pkgdown both render verbatim. -->
<!-- badges: start -->
[![R-CMD-check](https://github.com/statmodels7/numericals7/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/statmodels7/numericals7/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/statmodels7/numericals7/graph/badge.svg)](https://app.codecov.io/gh/statmodels7/numericals7)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(numericals7)
```

# numericals7

Every package of the [statmodels7](https://statmodels7.github.io) toolkit
needs the same numerical machinery -- higher-order derivatives of maps,
finite-difference stencils, quadrature, the combinatorics behind a chain
rule -- and before this package existed each had quietly grown its own copy:
the same set-partition enumeration was written twice, finite differences
three times. `{numericals7}` is that machinery written once, at the bottom of
the toolkit, where everything above can consume it.

## Installation

```{r eval=FALSE}
# install.packages("pak")
pak::pak("statmodels7/numericals7")
```

## Jets: a map written in ordinary R differentiates itself

A **jet** is a value carried together with every partial derivative up to
fourth order. Arithmetic on jets propagates those derivatives exactly, so an
expression that says nothing about derivatives carries all of them:

```{r}
lay <- jet_layout(2)
mu    <- jet_var(1, list(4.0, 1, 0, 0, 0), lay)
sigma <- jet_var(2, list(1.7, 1, 0, 0, 0), lay)

# the map from a Weibull's mean to its scale, as a reader would write it
scale <- mu / gamma(1 + 1 / sigma)
scale
```

The gradient, and every mixed derivative to fourth order, is exact to
machine precision -- there is no chain rule to transcribe and none to get
wrong. Comparisons and non-smooth functions are refused rather than
approximated: a branch taken on a jet would keep one side's derivatives and
report them as the whole expression's.

```{r}
# d scale / d sigma, by hand: mu * digamma(1 + 1/sigma) / (sigma^2 * gamma(...))
scale$d[[1]][2]
4.0 * digamma(1 + 1 / 1.7) / (1.7^2 * gamma(1 + 1 / 1.7))
```

## The enumerations behind a chain rule

A derivative of order four over several variables is a sum over combinatorial
objects, and an enumeration that exists in one copy cannot disagree with
itself:

```{r}
# the multi-indexes that key a derivative list (diagonal first at order 2)
tuple_indices(2, 2)

# the set partitions a Faa di Bruno formula sums over: 1, 2, 5, 15
lengths(lapply(1:4, set_partitions))

# the weak compositions of n into k parts -- the support of a multinomial
compositions(3, 2)
```

## One stencil library

`fd_weights()` builds the finite-difference weights for any derivative order
at any offsets from one Vandermonde system, `fd_offsets()` chooses the
offsets, `fd_step()` the step, and `fd_derivative()` assembles the three.
The toolkit's policy -- one stencil applied to the highest analytic order,
never a chain of first differences -- stays with the packages that enforce
it; this is the stencil itself, written once.

```{r}
# the classical five-point first-derivative weights, and a derivative
fd_weights(-2:2, 1L) * 12
fd_derivative(exp, 1, order = 3L, accuracy = 4L) - exp(1)
```

## Quadrature and series, vectorized over the parameters

A regression model gives every observation its own parameter value, so the
integrals a fit needs come in families: the same integrand at hundreds of
parameter rows. `quad_vec()` integrates all of them in one adaptive pass --
the integrand receives a matrix of points and a row index, so a thousand
parameter values cost matrix evaluations rather than a thousand adaptive
runs -- and `series_vec()` does the same for infinite sums. A row that
cannot reach the requested accuracy returns `NA` with a warning naming it.

```{r}
shp <- c(0.5, 2, 5, 50)
f <- function(x, i) x * dgamma(x, shape = shp[i], rate = 1)
quad_vec(f, 0, rep(Inf, 4))          # the means, one call

lam <- c(0.5, 4, 300)
series_vec(function(k, i) dpois(k, lam[i]), n = 3)   # masses sum to one
```

## What arrives next

The plan (`piano_numericals7.txt` in the toolkit repository) adds the special
functions the distributions carry with an overflow discipline attached --
Owen's T, the log-scale Mills ratio, the Bessel ratio and its inverse -- and
then the consumers above migrate package by package, the old implementations
serving as the independent reference until the comparison passes.
