---
title: "Numerical kernels"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Numerical kernels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(numericals7)
set.seed(1)
```

Almost every R package that computes a derivative it has no formula for writes
its own finite difference, and almost every one that sums a series writes its
own stopping rule. These are written as internal helpers, so other packages
cannot reuse them, and each is written to the accuracy that its author needed
at the time. A stencil is a fixed piece of arithmetic, and so are a quadrature
rule and the enumeration of the set partitions of four elements. This package
holds one copy of each for the packages above it, so that they all use the
same arithmetic and the same orderings.

Apart from the class `abs_smoother`, which is described in its own help pages,
the package consists of plain functions. This vignette describes five groups
of them and the limits of each.

## Derivatives with no closed form

A finite-difference stencil is the solution of a small Vandermonde system: given
offsets, the weights that reproduce a derivative of a given order exactly on
polynomials of as high a degree as the offsets allow. `fd_weights()` solves it
and returns the weights.

```{r}
fd_weights(-1:1, order = 1)     # the textbook central difference
fd_weights(-2:2, order = 1)     # five points, accurate two orders further
```

`fd_derivative()` applies one to a function. `accuracy` is the order of the
truncation error, and a higher accuracy costs only more evaluations of the
function:

```{r}
f  <- function(x) exp(-x^2 / 2)
d1 <- function(x) -x * exp(-x^2 / 2)          # the closed form, for comparison

sapply(c(2, 4, 6), function(a)
  abs(fd_derivative(f, 0.7, order = 1, accuracy = a) - d1(0.7)))
```

The step depends on the order $k$. The rounding error of a stencil for a
derivative of order $k$ grows like $\varepsilon h^{-k}$ as the step $h$
shrinks, and the truncation error at accuracy two shrinks like $h^2$. The two
balance at a step of $\varepsilon^{1/(k+2)}$ times the larger of one and the
magnitude of the point:

```{r}
data.frame(order = 1:4,
           step  = fd_step(1, 1:4),
           rule  = .Machine$double.eps^(1 / (1:4 + 2)))
```

A fourth derivative is therefore differenced at a step about four hundred
times that of a first derivative, and a fourth derivative computed at the step
of a first derivative is dominated by rounding error.

### Single stencils and nested differences

Throughout the toolkit a derivative of order $k$ is computed by **one** stencil
applied to the highest analytic order available, never by differencing a
difference. When each stage of a nested difference uses the step chosen for
its own order, the rounding errors multiply: three nested first differences of
$\sin$ at 0.7 are off by about $5 \times 10^{-2}$ in the third derivative, and
four are off by about $10^4$ in the fourth. Where a caller has an analytic
first derivative, the fourth derivative is one stencil of order three on it,
which is several digits more accurate than four nested first differences.

## Integrals and sums over a parameter vector

A modeling routine rarely needs a single integral. It needs the same integral
at every parameter setting in a vector, and computing them one at a time in a
loop pays R's call overhead once per setting. `quad_vec()` takes a vector of
endpoints and returns a vector of integrals, evaluating the integrand once for
every node of every panel of every row.

The integrand receives a matrix of evaluation points and an integer vector
that gives the parameter setting of each row, so the indexing is ordinary
recycling:

```{r}
a <- c(0.5, 1, 2, 5, 20)
got <- quad_vec(function(x, i) x^(a[i] - 1) * exp(-x),
                lower = rep(0, length(a)), upper = rep(Inf, length(a)))
rbind(quad_vec = got, gamma = gamma(a))
max(abs(got / gamma(a) - 1))
```

The number of rows comes from the **endpoints**, so scalar endpoints
give one integral however long the parameter vector is. Infinite endpoints are
mapped to finite ones by a rational transform whose Jacobian multiplies the
integrand, and rows of different kinds may share one call.

Adaptivity is judged per row on the **sum** of that row's panel errors, as
QUADPACK judges it. A budget allocated panel by panel cannot meet an integrable
endpoint singularity at any depth, the error there being concentrated in the
innermost panel however deep the bisection goes.

`series_vec()` is the same idea for a sum. Its stopping rule estimates the
neglected tail from the decay of the terms at the end of the last block, and a
small block sum is not enough: a block sum and a last term that are both small
occur long **before** the mode of a hump-shaped term.

```{r}
# The first 64 terms of E[Y^2] for a Poisson mean of 300.
sum((0:63)^2 * dpois(0:63, 300))   # the block sum: tiny
63^2 * dpois(63, 300)              # the last term: tiny
300 + 300^2                        # what the sum actually is
```

The terms of that block are rising, so their decay ratio is above one and the
estimated tail is not small. The row continues until the terms decay and the
geometric continuation of that decay, doubled, fits the budget:

```{r}
lam <- c(0.5, 3, 40, 300)
got <- series_vec(function(k, i) k^2 * dpois(k, lam[i]), n = length(lam))
max(abs(got / (lam + lam^2) - 1))
```

### Rows that do not converge

A row whose panels still exceed their budget at `max_depth`, or that holds
`max_panels` panels, returns `NA`, and one warning lists the first eight such
rows:

```{r}
suppressWarnings(
  quad_vec(function(x, i) 1 / x, lower = 0, upper = 1, max_depth = 6)
)
```

The integral diverges, so no accuracy can be reached, and the value `NA`
indicates that the budget was not met.

## Functions evaluated on a representable scale

Several quantities that a likelihood needs are finite and ordinary while the
expression normally used to compute them overflows or underflows. Each of these is written on
the scale where it stays representable.

The inverse Mills ratio $R(t) = \phi(t)/\Phi(t)$ is asymptotic to $-t$, so it is
about 100 at $t = -100$. Both of its factors underflow well before that:

```{r}
t <- c(-5, -38, -100, -400)
m <- mills_ratio(t)
data.frame(t, naive = dnorm(t) / pnorm(t), mills = m$r, asymptote = -t)
```

It also returns the derivative, which lies in $(-1, 0)$, so a caller does not
compute it from `r` a second time. The modified Bessel function of
the first kind overflows at an argument no larger than a fitted concentration:

```{r}
x <- c(10, 500, 800, 5000)
data.frame(x, naive = besselI(x, 0), log_bessel = log_bessel_i(x, 0))
```

and the ratio $I_1/I_0$, which a von Mises score needs, is finite everywhere
while the ratio of the two scaled functions is not:

```{r}
k <- c(1, 1e4, 1e5, 1e6)
data.frame(
  kappa = k,
  naive = besselI(k, 1, expon.scaled = TRUE) /
          besselI(k, 0, expon.scaled = TRUE),
  ratio = bessel_i_ratio(k)
)
```

Owen's T is included because the skew-normal distribution function is written
in terms of it, and Owen's T is a bounded one-dimensional integral over the
interval from 0 to $a$:

```{r}
h <- 0.7; a <- 1.3
owen_t(h, a)
integrate(function(x) exp(-h^2 * (1 + x^2) / 2) / (1 + x^2), 0, a)$value / (2 * pi)
```

## The enumerations

A higher-order chain rule is a sum over set partitions, a symmetric derivative
array is indexed by non-decreasing tuples, and the support of a multinomial is
the weak compositions. All three are fixed combinatorial objects, and they are
collected here because several packages need them.

```{r}
length(set_partitions(4))          # the Bell number B4
choose(2 + 3 - 1, 3)               # tuples of order 3 on 2 parameters
length(tuple_indices(2, order = 3))
```

`set_partitions(n)` returns each partition as a list of blocks, from the one
block containing everything to the $n$ singletons. `tuple_indices(d, order)`
returns non-decreasing index vectors, which is exactly the set of distinct
components of a symmetric derivative array:

```{r}
sapply(tuple_indices(2, order = 3), paste, collapse = "")
```

`compositions(n, k)` returns the ways of writing $n$ as an ordered sum of $k$
non-negative parts, one per row:

```{r}
compositions(3, 2)
rowSums(compositions(3, 2))
```

## Threads

`n_threads()` carries the thread policy of the whole toolkit. It is an object
passed as an argument from `statmod()` and `fit_distrib()` down to the kernels,
so no package reads a setting that lives in another. `local_threads()` sizes the
RcppParallel worker pool for the duration of a fit and restores the previous
setting on exit. The object lives at the root of the dependency graph because
the packages holding the kernels cannot import the one at the top.

```{r}
th <- n_threads(threads = 4, workers = 2)
c(threads = thread_count(th), workers = worker_count(th))
```

The guarantee on its page is that a kernel's result does not depend on the count
bit for bit:

```{r}
x <- seq(0.5, 50, length.out = 2000)
identical(log_bessel_i(x, 0, threads = 1), log_bessel_i(x, 0, threads = 4))
```

`log_bessel_k()` takes no count. Its hybrid branch calls R's own `besselK`,
which can raise a warning, and a warning raised from a worker thread terminates
the process.

## Summary

* **Stencils.** `fd_offsets()` and `fd_weights()` give the arithmetic,
  `fd_derivative()` applies it, `fd_step()` gives the step that balances
  rounding against truncation. One stencil on the highest analytic order
  available, never a difference of a difference.
* **Batched engines.** `quad_vec()` and `series_vec()` are vectorized over the
  parameter and return `NA` for a row whose budget was not met. `quad_vec()`
  takes its row count from the endpoints and judges convergence on the sum of
  a row's panel errors; `series_vec()` takes it from `n` and retires a row
  when the last block and the estimated tail both fit the budget.
* **Special functions.** `mills_ratio()`, `owen_t()`, `bessel_i_ratio()`,
  `log_bessel_i()` and `log_bessel_k()`, each written on the scale where it
  stays representable. `mills_ratio()` returns its derivative with the value;
  the derivatives of the Bessel functions are separate functions
  (`bessel_i_ratio_d1()` to `_d4()`, `log_bessel_i_derivs()`,
  `log_bessel_k_derivs()`).
* **Enumerations.** `set_partitions()`, `tuple_indices()` and `compositions()`,
  one copy each for the packages of the toolkit.
* **Threads.** `n_threads()` is passed as an argument, and a kernel's answer is
  bit-identical at any count.
